Bayesian hidden Markov model analysis of single-molecule force spectroscopy:\\ Characterizing kinetics under measurement uncertainty

### 2015-02-18 manuscripts/bayesian-hmm-method/bhmm.tex
wrote T-matrix sampling description in manuscript appendix

\usepackage{amsmath}
\usepackage{amssymb}

@@

The transition matrix is updated using the reversible transition matrix sampling scheme of No\'{e}~\cite{noe:jcp:2008:transition-matrix-sampling,chodera-noe:jcp:2011:observable-sampling}.
Here, an adjusted count matrix $\bfm{C} \equiv (c_{ij})$ is computed using the updated hidden state sequence $\bfm{S}'$,
\begin{eqnarray}
c_{ij} &=& b_{ij} +\sum_{t=1}^{L} \delta_{is_{t-1}} \delta_{js_{t}} ,
\end{eqnarray}
where the Kronecker $\delta_{ij} = 1$ if $i = j$ and zero otherwise, and $\bfm{B} \equiv (b_{ij})$ is a matrix of prior pseudocounts, which we take to be zero following the work of No\'{e} et al.~\cite{noe:pnas:2009:ww-domain}.
Using the adjusted count matrix $\bfm{C}$, a Metropolis-Hastings Monte Carlo procedure~\cite{metropolis-hastings} is used to update the matrix and produce a new sample from $\Pr(\bfm{T}' \mid \bfm{S}')$.
Two move types are attempted, selected with equal probability, and 1000 trial moves are attempted to generate a new sample $\bfm{T}'$ that is approximately uncorrelated from the previous $\bfm{T}$.
Prior to starting the Monte Carlo procedure, the vector of equilibrium probabilities for all states $\bfv{\pi}$ is computed according to 
\begin{eqnarray}
\bfm{T}^\T \bfv{\pi} = \bfv{\pi} .
\end{eqnarray}

The first move type is a reversible element shift.
A pair of states $(i,j)$, $i \ne j$, are selected with uniform probability, and a random number $\Delta$ is selected uniformly over the interval,
\begin{equation*}
\Delta \in[\max(-T_{ii},-\frac{\pi_{j}}{\pi_{i}}T_{jj}),T_{ij}] .
\end{equation*}
The changed elements in the proposed transition matrix $\bfm{T}'$ are then given by:
\begin{align*}
T'_{ij} & =T_{ij}-\Delta \:\: ; \:\: T'_{ji}  =T_{ji}-\frac{\pi_{i}}{\pi_{j}}\Delta \\
T'_{ii} & =T_{ii}+\Delta \:\: ; \:\: T'_{jj}  =T_{jj}+\frac{\pi_{i}}{\pi_{j}}\Delta  .
\end{align*}
This move is accepted with probability
\begin{eqnarray}
\lefteqn{P_\mathrm{accept}(\bfm{T}' | \bfm{T} ) = \mathrm{min}\left\{1, \sqrt{\frac{(T'_{ij})^{2}+(T'_{ji})^{2}}{(T_{ij})^{2}+(T_{ji})^{2}}} \right. }\\
&\times& \left. \left(\frac{T'_{ii}}{T_{ii}}\right)^{c_{ii}}\left(\frac{T'_{ij}}{T_{ij}}\right)^{c_{ij}} \left(\frac{T'_{jj}}{T_{jj}}\right)^{c_{jj}}\left(\frac{T'_{ji}}{T_{ji}}\right)^{c_{ji}} \right\} . \nonumber
\end{eqnarray}
This move will leave the vector of stationary probabilities $\bfv{\pi}$ unchanged.

The second move type is a row shift.
A row $i$ of $\bfm{T}$ is selected with uniform probability, and a random number $\eta$ chosen uniformly over the interval 
\begin{equation*}
\eta\in\left[0,\frac{1}{1-T_{ii}}\right]
\end{equation*}
and used to update row $i$ of $\bfm{T}$ according to
\begin{eqnarray}
T'_{ij} &=& \begin{cases}
\eta T_{ij} & j = 1, \ldots, M, \:\: j \ne i\\
\eta (T_{ii} - 1) + 1 & j = i
\end{cases}
\end{eqnarray}
This move is accepted with probability
\begin{eqnarray}
P_\mathrm{accept}(\bfm{T}' | \bfm{T}) &=& \mathrm{min} \left\{1, \eta^{(M-2)} \eta^{(c_{i*}-c_{ii})}\left(\frac{1-\eta(1-T_{ii})}{T_{ii}}\right)^{c_{ii}} \right\} . \nonumber\\
\end{eqnarray}
The row shift operation will change the stationary distribution of $\bfv{\pi}'$, but it may be efficiently updated with
\begin{align*}
\pi'_{i} & =\frac{\pi_{i}}{\pi_{i}+\eta(1-\pi_{i})} \:\: ; \:\: \pi'_{j} =\frac{\eta \, \pi_{j}}{\pi_{i}+\eta(1-\pi_{i})}.
\end{align*}
Since this update scheme is incremental, it will accumulate numerical errors over time that cause the updated $\bfv{\pi}$ to drift away from the stationary distribution of the current transition matrix. 
To avoid this, $\bfv{\pi}$ is recomputed from the current sample of the transition matrix in regular intervals (here, every 100 sampling steps).

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[]{algorithm2e}

@@

%%%%%%%%%%%%%%

\color{magenta}

Here we employ the reversible transition matrix sampling algorithm derived in [{\bf FN: reference to reversible T-matrix estimation preprint when available}]. We first define a coordinate transform to unconditional transition probabilities:
\begin{equation}
  \label{eq:transformation_T_X}
  x_{ij} = \pi_{i} T_{ij}
\end{equation}
and sample from their posterior 
\begin{equation}
  \label{eq:X_posterior}
   p(\mathbf{X} \mid \mathbf{C} ) = p(\mathbf{C} \mid \mathbf{X}) p(\mathbf{X})
\end{equation}
using the prior
\begin{equation}
  \label{eq:X_prior}
   p(\mathbf{X}) = \prod_{i \ge j} x_{ij}^{-1}.
\end{equation}
This choice of prior ensures that the maximum likelihood estimator, i.e. the maximum of $p(\mathbf{C} \mid \mathbf{X})$ coincides with
the mean of the posterior $p(\mathbf{X} \mid \mathbf{C} )$. This behavior is desirable in order to have uncertainty intervals envelop the
maximum likelihood estimator. Without a suitable choice of prior this property is lost, as described in~\cite{noe:pnas:2009:ww-domain}.

It turns out that the posterior (\ref{eq:X_posterior}) can be conveniently sampled by considering yet another set of
variables $\mathbf{V}=[v_{ij}]$, which are related to $\mathbf{X}$ by the scaling $x_{ij} = v_{ij} / \sum_{k,l} v_{kl}$. Thus, we can still recover 
transition matrices from the $v_{ij}$ variables by row normalization:
\begin{equation}
  \label{eq:T_from_V}
  T_{ij} = \frac{v_{ij}}{\sum_k v_{ik}}
\end{equation}

Using the $v_{ij}$ variables, the sampling of posterior (\ref{eq:X_posterior}) is implemented by iterating the following Gibbs sampling
procedure. Initially, when given an initial transition matrix $\mathbf{T}^{(0)}$, we set 
\begin{equation} 
	v_{ij}^{(0)} = x_{ij}^{(0)} = \pi_{i} T_{ij}^{(0)}
\end{equation} 
When starting from scratch we initialize 
\begin{equation} 
	v_{ij}^{(0)} = (c_ij + c_ji) / 2 \sum_{k,l} c_{kl}.
\end{equation} 
In each Gibbs sampling cycle we iterate through all $i \ge j$. If $c_{ij} + c_{ji} = 0$, the conditional probability density is concentrated at
$v_{ij} = 0$ and we do not perform any action. If  $c_{ij} + c_{ji} > 0$, one of the following two steps is executed for either diagonal or off-diagonal
elements:

%% diagonals
{\bf Case 1: diagonals}: If $i = j$, we sample $v_{ii}^{(k+1)}$ according to the conditional
\begin{eqnarray}
  \mathbb{P}(v_{ii} | \{v_{kl}\}_{k \geq l} \backslash \{v_{ii}\}, \mathbf{C}) 
							& \propto & v_{ii}^{c_{ii}-1} (v_i - v_{ii})^{-c_{i}}. \\
							& \propto & T_{ii}^{c_{ii}-1} (1 - T_{ii})^{c_{i}-c_{ii}-1}.
  \label{eq:conditional_pkk}
\end{eqnarray}
where we have introduced the row sums
\begin{equation}
   v_i = \sum_j v_{ij}.
\end{equation}
The conditional (\ref{eq:conditional_pkk}) can be sampled by drawing Beta-distributed variables 
\begin{equation}
	T_{ii}\sim \mathrm{Beta}(\alpha,\beta)
	\label{eq:betadist}
\end{equation}
with parameters
$\alpha=c_{ii}$ and $\beta=c_{i}-c_{ii}$ (see \cite{Devroye_Springer86_NonuniformRandom}). The new element $v_{ii}^{(k+1)}$
can be obtained from the sampled $T_{ii}$ by
\begin{equation}
  v_{ii}^{(k+1)}=\frac{T_{ii}}{1-T_{ii}} (v_i^{(k)} - v_{ii}^{(k)}).
  \label{eq:diagonal_transformation}
\end{equation}

% off-diagonals.
{\bf Case 2: off-diagonals}: If $i \neq j$, we have to sample a conditional density which is not easily tractable. Therefore,
we approximate the conditional density as follows:
\begin{equation}
  \begin{aligned}
&  v_{ij}^{(c_{ij}+c_{ji}-1)} (v_i^{(k)} - v_{ij}^{(k)} + v_{ij})^{-c_i} (v_j^{(k)} - v_{ji}^{(k)} + v_{ij})^{-c_j}  \\ 
& \approx q(v_{ij} \mid \alpha, \beta, f_0)
  \end{aligned}
  \label{eq:conditional_vkl_f}
\end{equation}
where employ the approximate function
\begin{equation}
	q(v_{ij} \mid \alpha, \beta, f_0)         = (v_{ij})^{-1} \exp \left( \alpha \log v_{ij} - \beta v_{ij} + f_0 \right)
	\label{eq:approximation_conditional}
\end{equation}
which is, up to a constant factor, a Gamma distribution with parameters $\alpha$, $\beta$ which can be
efficiently sampled \cite{Devroye_Springer86_NonuniformRandom}. The parameters are given by
\begin{subequations}
  \begin{align}
    \alpha & = -f''(\bar{v}_{ij}) \bar{v}_{ij}^2 \label{eq:alpha}\\
    \beta & = -f''(\bar{v}_{ij}) \bar{v}_{ij} \label{eq:beta} \\
    f_0 & = f(\bar{v}_{ij}) +  f''(\bar{v}_{ij}) \bar{v}_{ij}^2 (\log \bar{v}_{ij} -1) \label{eq:f0},
  \end{align}
\end{subequations}
with the maximum point $\bar{v}_{ij}$ as the positive root of a quadratic equation:
\begin{equation}
  \label{eq:maximum_point}
  \bar{v}_{ij} = \frac{-b + \sqrt{b^2-4ac}}{2a}
\end{equation}
and $a$, $b$, $c$ given by:
\begin{subequations}
  \begin{align}
    a &=c_{i}+ c_{j}- c_{ij}- c_{ji} \label{eq:a} \\
    b &= c_{i} (v_i^{(k)} - v_{ij}^{(k)}) + c_{j} (v_j^{(k)} - v_{ji}^{(k)}) \nonumber \\
       & - (c_{ij}+c_{ji}) (v_i^{(k)} - v_{ij}^{(k)} + v_j^{(k)} - v_{ji}^{(k)}) \label{eq:b} \\
    c &= -(c_{ij}+c_{ji}) (v_i^{(k)} - v_{ij}^{(k)}) (v_j^{(k)} - v_{ji}^{(k)}) \label{eq:c}
  \end{align}
\end{subequations}
The approximate conditional $q$ in \eqref{eq:approximation_conditional} is
not necessarily a bounding envelope for the true conditional
\eqref{eq:conditional_vkl_f} so that instead of rejection sampling we
use the approximation $q$ as a proposal density in a Metropolis sampling
step. The acceptance probability for the step is then given as $\min
\{1, p_{acc}\}$ with
\begin{equation}
  \label{eq:pacc}
  p_{acc}=\frac{ q(v_{ij}^{(k)}) v_{ij}^{(k)} \exp f(v_{ij}^{(k+1)})}{ q(v_{ij}^{(k+1)}) v_{ij}^{(k+1)}  \exp f(v_{ij}^{(k)})}.
\end{equation}

For each element sampled, symmetry of the $\mathbf{V}$ matrix is enforced by setting $v_{lk} = v_{kl}$. After each
cycle of the Gibbs sampling procedure, we rescale $\mathbf{V}$ by
\begin{equation}
  v_{ij}^{(k+1)} \leftarrow \frac{v_{ij}^{(k+1)}}{\sum_{l,m} v_{lm}^{(k+1)}}
\end{equation}

In brief, we can summarize the above procedure by the algorithm:
\begin{algorithm}
  \DontPrintSemicolon
  \SetAlgoNoLine
  \SetInd{0.5em}{1.0em}
  \KwIn{$C$, $V^{(k)}$}
  \KwOut{$V^{(k+1)}$}
  \For{$i=1,\dotsc,n$}{
    \For{$j=1,\dotsc, i$}{
      \If{$c_{ij}+c_{ji}>0$}{
        \eIf{$i=j$}{
          % $\alpha=c_{kk}$, $\beta=c_{k}-c_{kk}$\;
          % $p_{kk}=\frac{v^{(j)}_{kk}}{v^{(j)}_{k,-k}+v^{(j)}_{kk}}$\;
          % $p'_{kk} \sim \text{Beta}(\alpha, \beta)$\;
          % $v_{kk}^{(j+1)}=\frac{p'_{kk}}{1-p'_{kk}}v_{k, -k}^{(j)}$\;
          Sample $v^{(k+1)}_{ii}$ according to
          \eqref{eq:conditional_pkk}, \eqref{eq:transformation_p_v}\;
        }{ Obtain parameters $a$, $b$, $c$ from \eqref{eq:a},
          \eqref{eq:b}, \eqref{eq:c}\; Compute $\bar{v}_{ij}$
          according to \eqref{eq:maximum_point}\; Compute parameters
          $\alpha$, $\beta$, $f_0$ using \eqref{eq:alpha},
          \eqref{eq:beta},\eqref{eq:f0}\; Sample $v'_{ij}$ according
          to \eqref{eq:approximation_conditional}\; Compute $p_{acc}$
          from \eqref{eq:pacc}\; $U$ from $[0, 1]$ uniformly
          distributed\; \eIf{$U<\min\{1, p_{acc}\}$}{
            $v_{ij}^{(k+1)}=v'_{ij}$\; }{
            $v_{ij}^{(k+1)}=v^{(k)}_{ij}$\; }
          $v_{ji}^{(j+1)}=v_{ij}^{(k+1)}$\; } } } }
\caption{Reversible sampling algorithm \label{alg:reversible_sampler}}
\end{algorithm}

\color{black}

%The transition matrix is updated using the reversible transition matrix sampling scheme of No\'{e}~\cite{noe:jcp:2008:transition-matrix-sampling,chodera-noe:jcp:2011:observable-sampling}.
%Here, an adjusted count matrix $\bfm{C} \equiv (c_{ij})$ is computed using the updated hidden state sequence $\bfm{S}'$,
%\begin{eqnarray}
%c_{ij} &=& b_{ij} +\sum_{t=1}^{L} \delta_{is_{t-1}} \delta_{js_{t}} ,
%\end{eqnarray}
%where the Kronecker $\delta_{ij} = 1$ if $i = j$ and zero otherwise, and $\bfm{B} \equiv (b_{ij})$ is a matrix of prior pseudocounts, which we take to be zero following the work of No\'{e} et al.~\cite{noe:pnas:2009:ww-domain}.
%Using the adjusted count matrix $\bfm{C}$, a Metropolis-Hastings Monte Carlo procedure~\cite{metropolis-hastings} is used to update the matrix and produce a new sample from $\Pr(\bfm{T}' \mid \bfm{S}')$.
%Two move types are attempted, selected with equal probability, and 1000 trial moves are attempted to generate a new sample $\bfm{T}'$ that is approximately uncorrelated from the previous $\bfm{T}$.
%Prior to starting the Monte Carlo procedure, the vector of equilibrium probabilities for all states $\bfv{\pi}$ is computed according to 
%\begin{eqnarray}
%\bfm{T}^\T \bfv{\pi} = \bfv{\pi} .
%\end{eqnarray}
%
%The first move type is a reversible element shift.
%A pair of states $(i,j)$, $i \ne j$, are selected with uniform probability, and a random number $\Delta$ is selected uniformly over the interval,
%\begin{equation*}
%\Delta \in[\max(-T_{ii},-\frac{\pi_{j}}{\pi_{i}}T_{jj}),T_{ij}] .
%\end{equation*}
%The changed elements in the proposed transition matrix $\bfm{T}'$ are then given by:
%\begin{align*}
%T'_{ij} & =T_{ij}-\Delta \:\: ; \:\: T'_{ji}  =T_{ji}-\frac{\pi_{i}}{\pi_{j}}\Delta \\
%T'_{ii} & =T_{ii}+\Delta \:\: ; \:\: T'_{jj}  =T_{jj}+\frac{\pi_{i}}{\pi_{j}}\Delta  .
%\end{align*}
%This move is accepted with probability
%\begin{eqnarray}
%\lefteqn{P_\mathrm{accept}(\bfm{T}' | \bfm{T} ) = \mathrm{min}\left\{1, \sqrt{\frac{(T'_{ij})^{2}+(T'_{ji})^{2}}{(T_{ij})^{2}+(T_{ji})^{2}}} \right. }\\
%&\times& \left. \left(\frac{T'_{ii}}{T_{ii}}\right)^{c_{ii}}\left(\frac{T'_{ij}}{T_{ij}}\right)^{c_{ij}} \left(\frac{T'_{jj}}{T_{jj}}\right)^{c_{jj}}\left(\frac{T'_{ji}}{T_{ji}}\right)^{c_{ji}} \right\} . \nonumber
%\end{eqnarray}
%This move will leave the vector of stationary probabilities $\bfv{\pi}$ unchanged.
%
%The second move type is a row shift.
%A row $i$ of $\bfm{T}$ is selected with uniform probability, and a random number $\eta$ chosen uniformly over the interval 
%\begin{equation*}
%\eta\in\left[0,\frac{1}{1-T_{ii}}\right]
%\end{equation*}
%and used to update row $i$ of $\bfm{T}$ according to
%\begin{eqnarray}
%T'_{ij} &=& \begin{cases}
%\eta T_{ij} & j = 1, \ldots, M, \:\: j \ne i\\
%\eta (T_{ii} - 1) + 1 & j = i
%\end{cases}
%\end{eqnarray}
%This move is accepted with probability
%\begin{eqnarray}
%P_\mathrm{accept}(\bfm{T}' | \bfm{T}) &=& \mathrm{min} \left\{1, \eta^{(M-2)} \eta^{(c_{i*}-c_{ii})}\left(\frac{1-\eta(1-T_{ii})}{T_{ii}}\right)^{c_{ii}} \right\} . \nonumber\\
%\end{eqnarray}
%The row shift operation will change the stationary distribution of $\bfv{\pi}'$, but it may be efficiently updated with
%\begin{align*}
%\pi'_{i} & =\frac{\pi_{i}}{\pi_{i}+\eta(1-\pi_{i})} \:\: ; \:\: \pi'_{j} =\frac{\eta \, \pi_{j}}{\pi_{i}+\eta(1-\pi_{i})}.
%\end{align*}
%Since this update scheme is incremental, it will accumulate numerical errors over time that cause the updated $\bfv{\pi}$ to drift away from the stationary distribution of the current transition matrix. 
%To avoid this, $\bfv{\pi}$ is recomputed from the current sample of the transition matrix in regular intervals (here, every 100 sampling steps).
